A Linear Solution of the Linearity Challenge

### 2025-12-24 paper/Context.lagda.tex
revised typing contexts

We are going to adopt a nameless representation of channels whereby a channel is
identified by the position of its type in a typing context. This representation
is akin to using De Bruijn indices except that the index, instead of being
represented explicitly by a natural number, is computable from the proof that
the type of the channel belongs to the typing context.
%
Typing contexts are represented using lists of types, where the (polymorphic)
\AgdaDatatype{List} type and its constructors \AgdaInductiveConstructor{[]} and
\AgdaInductiveConstructor{\_∷\_} are defined in the module
\AgdaModule{Data.List} of Agda's standard library. We will keep using
$\ContextC$, $\ContextD$ and $\ContextE$ to range over typing contexts, even
though in the Agda formalization they are lists instead of finite maps.

@@

Note that, when specifying a splitting $\Splitting\ContextC{[A]}\ContextE$, it
may be convenient to use \AgdaFunction{≫} as a shortcut for a sequence of
applications of \AgdaInductiveConstructor{>} after the $A$ type has been
reached. For instance, \AgdaFunction{splitting-example₁} can be written
equivalently and in a more compact way as shown below. More usages of
\AgdaFunction{≫} will be given in \Cref{sec:examples}.

@@

In the following we will make extensive use of predicates over contexts. For
this reason, it is worth recalling a number of definitions from the
\AgdaModule{Relation.Unary} module in Agda's standard library in
\Cref{tab:predicates}.

Context splitting is going to be pervasive in the subsequent development, to the
point that it contributes substantially to the complexity of the formalization.
For this reason, we invest some effort in defining a few abstractions that allow
us to occasionally hide context splittings and enhance code readability.
%
Inspired by separation logic~\cite{OHearnPym99} and
following~\citet{RouvoetPoulsenKrebbersVisser20}, we define the \emph{separating
conjunction} $P \sand Q$ of two predicates $P$ and $Q$ over contexts:
\begin{code}
data _∗_ (P Q : Pred Context _) (Γ : Context) : Set where
  _⟨_⟩_ : ∀{Δ Θ} → P Δ → Γ ≃ Δ + Θ → Q Θ → (P ∗ Q) Γ
\end{code}

The predicate $P \sand Q$ holds for those contexts $\ContextC$ that can be split
into $\ContextD$ and $\ContextE$ where $P$ holds for $\ContextD$ and $Q$ holds
for $\ContextE$.
%
The constructor \AgdaInductiveConstructor{\_⟨\_⟩\_} has three explicit arguments
witnessing the splitting $\Splitting\ContextC\ContextD\ContextE$ along with
$P~\ContextD$ and $Q~\ContextE$.\Luca{Nei related fare presente che queste
definizioni seguono sostanzialmente \citet{RouvoetPoulsenKrebbersVisser20} ed
evidenziare le differenze.}

>>>>

We are going to adopt a nameless representation of channels hence typing
contexts are represented using lists of types, where the (polymorphic)
\AgdaDatatype{List} type and its constructors \AgdaInductiveConstructor{[]} and
\AgdaInductiveConstructor{\_∷\_} are defined in the module
\AgdaModule{Data.List} of Agda's standard library. We will keep using
$\ContextC$, $\ContextD$ and $\ContextE$ to range over typing contexts as we
have done in \Cref{sec:typing-rules}, even though in the Agda formalization they
are lists instead of finite maps.

@@

Note that, when specifying a splitting $\Splitting\ContextC{[A]}\ContextE$, it
may be convenient to use \AgdaFunction{≫} as a shortcut for a sequence of
applications of \AgdaInductiveConstructor{>} after the $A$ type has been
reached. For instance, \AgdaFunction{splitting-example₁} can be written
equivalently and in a more compact way as shown below. More usages of
\AgdaFunction{≫} will be provided in \Cref{sec:examples}.

@@

Hereafter we will make extensive use of predicates over contexts. For this
reason, it is worth recalling a number of definitions from the
\AgdaModule{Relation.Unary} module in Agda's standard library in
\Cref{tab:predicates}.

Context splitting is going to be pervasive in the subsequent development, to the
point that it contributes substantially to the complexity of the formalization.
For this reason, we invest some effort in defining a few abstractions that allow
us to hide context splittings and enhance code readability.
%
Inspired by separation logic~\cite{OHearnPym99} and
following~\citet{RouvoetPoulsenKrebbersVisser20}, we define the \emph{separating
conjunction} $P \sand Q$ of two predicates $P$ and $Q$ over contexts:
\begin{code}
data _∗_ (P Q : Pred Context _) (Γ : Context) : Set where
  _⟨_⟩_ : ∀{Δ Θ} → P Δ → Γ ≃ Δ + Θ → Q Θ → (P ∗ Q) Γ
\end{code}

The predicate $P \sand Q$ holds for those contexts $\ContextC$ that can be split
into $\ContextD$ and $\ContextE$ where $P$ holds for $\ContextD$ and $Q$ holds
for $\ContextE$.
%
The constructor \AgdaInductiveConstructor{\_⟨\_⟩\_} has three explicit arguments
witnessing the splitting $\Splitting\ContextC\ContextD\ContextE$ along with
$P~\ContextD$ and $Q~\ContextE$.
%
The use of the metavariables $P$ and $Q$ for denoting predicates over contexts
is not without reason. As we will see shortly, in our formalization processes
are indeed predicates over typing contexts, albeit not the only ones.
